\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-07-31 docs/parco13/paper.tex
Go through paper up to and including sec 3

\begin{abstract}
We present a new hybrid paradigm for parallel adaptive mesh refinement (AMR)
that combines the scalability and lightweight architecture of tree-based AMR
with the computational efficiency of patch-based solvers for hyperbolic
conservation laws.  The key idea is to interpret each leaf of the AMR hierarchy
as one uniform compute patch in $\sR^d$ with $m^d$ degrees of freedom, where $m$ is
customarily between 8 and 32.  Thus, computation on each patch can be optimized
for speed, while we inherit the flexibility of adaptive meshes.  In our work we
choose to work with the \pforest AMR library since it allows us to compose the mesh
from multiple mapped octrees and enables the cubed sphere and other
nontrivial multiblock geometries.
\end{abstract}

@@

In this paper we describe our design for the coupling of forest-of-octree AMR
with \clawpack at the leaf level.  We comment on challenges that arise in
enabling multiblock geometries and efficient parallelism and conclude with a range
of numerical examples that demonstrate the conceptual improvements in relation
to other approaches.

@@

\alert{I have swapped sections 3. and 4.  Please check/smooth the text.}

For hyperbolic problems, we integrate the solution on a single uniform patch,
containing $m^d$ cells, using the wave propagation algorithm described by R. J.
LeVeque \cite{LeVeque:1997eg} and implemented in \clawpack \cite{le:2002,
clawpack}.  We assume a single degree of freedom per cell and reconstruct a
piecewise constant solution to obtain left and right states at cell edges.  At
each edge, we solve Riemann problems to obtain left and right going waves
propagating at speeds determined from the solution to the Riemann problem.  For
scalar advection, the speed of each wave is simply the local advection speed at
the cell interface.  For non-linear problems and systems, an approximate
Riemann solver, such as a Roe solver \cite{roesolver}, is typically used.  Since
much of the physics of an application can be contained in the Riemann solver,
\forestclaw adopts \clawpack's interface to Riemann solvers effectively allowing
problems solved with \clawpack to be solvable in \forestclaw.  Wave limiters
are used to supress spurious oscillations resulting from truncation error in a typical
second order method.
\comment{DAC: Limiters formally reduce the order of accuracy.}
\ignore{In order to
achieve second order accuracy wave limiters are used to suppress spurious
oscillations at sharp gradients in the flow.}

Data exchanges between neighboring patches are done via layers of
ghost cells extending the dimensions along the edges of each patch.
The interior edge values of a given patch overlap the ghost cell
region of a neighboring patch.  For the second order wave propagation
algorithm, two layers of ghost cells are sufficient.  This implies that
one layer of ghost patches is sufficient for $m \ge 4$.  Neighboring
patches at the same level of refinement simply copy their interior
edge values into a neighbors' ghost cells.  Neighboring fine grid
patches average their interior edge data to a coarser neighbor's ghost
cell values.  And neighboring coarse grid patches interpolate data
from their interior edge cells to their fine grid neighbor's ghost
cell values.  To avoid loss of conservation and the creation of
spurious extrema, we use a standard conservative, limited
interpolation scheme to interpolate values from the coarse grid to
fine grid coarse cells \cite{amrclaw, chombo}.  When sub-cycling, time accurate
data between coarse grids is used to fill in ghost cells for fine grids.
As mentioned in the previous section, this procedure can be extended
transparently to distributed parallelism by defining an abstract exchange
routine for ghost patch data.

@@

\begin{figure}
\begin{center}
\includegraphics[width=.4\columnwidth]{../figs_misc/forest_stacked}
\hspace{.05\columnwidth}
\begin{tikzpicture}[every node/.style={anchor=south west,inner sep=0pt},
                    x=1mm, y=1mm]
  \node at (0,0) {
    \includegraphics[width=.4\columnwidth]{../figs_misc/cubed3}
  };
  \node [fill=white, inner sep=2pt] at (28, 10) {1};
  \node [fill=white, inner sep=2pt] at (5, 27) {2};
  \node [fill=white, inner sep=2pt] at (41, 27) {2};
  \node [fill=white, inner sep=2pt] at (28, 36) {3};
\end{tikzpicture}
%\includegraphics[width=.4\columnwidth]{../figs_misc/cubed3}
\end{center}
\caption{Left: Forest of two quadtrees, partitioned among three MPI processes.
Each quadtree has its own coordinate orientation.  The situation in 3D
(octrees) is analogous.  Right: The leaves in a forest of six quadtrees that
serves as the computational domain for the cubed sphere.  An adhoc refinement
pattern has been 2:1 balanced in terms of neighbor sizes and partitioned
between five MPI processes (the three middle ones out of $0 \ldots 4$
are shown from black to white).}
\label{fig:cubed3}
\end{figure}

>>>>

\begin{abstract}
We present a new hybrid paradigm for parallel adaptive mesh refinement (AMR)
that combines the scalability and lightweight architecture of tree-based AMR
with the computational efficiency of patch-based solvers for hyperbolic
conservation laws.  The key idea is to interpret each leaf of the AMR hierarchy
as one uniform compute patch in $\sR^d$ with $m^d$ degrees of freedom, where $m$ is
customarily between 8 and 32.  Thus, computation on each patch can be optimized
for speed, while we inherit the flexibility of adaptive meshes.  In our work we
choose to integrate with the \pforest AMR library since it allows us to compose
the mesh from multiple mapped octrees and enables the cubed sphere and other
nontrivial multiblock geometries.  We describe aspects of the parallel
implementation and close with scalings for both MPI-only and OpenMP/MPI hybrid
runs, where the largest MPI-only run uses 4,096 CPU cores.
\end{abstract}

@@

In this paper we describe our design for the coupling of forest-of-octrees AMR
with \clawpack at the leaf level.  We comment on challenges that arise in
enabling multiblock geometries and efficient parallelism and conclude with a range
of numerical examples that demonstrate the conceptual advantages.
% improvements in relation to other approaches.

@@

For hyperbolic problems, we integrate the solution on a single uniform patch,
containing $m^d$ cells, using the wave propagation algorithm described by R. J.
LeVeque \cite{LeVeque:1997eg} and implemented in \clawpack \cite{le:2002,
clawpack}.  We assume a single degree of freedom per cell and reconstruct a
piecewise constant solution to obtain left and right states at cell edges.  At
each edge, we solve Riemann problems to obtain left and right going waves
propagating at speeds determined from the solution to the Riemann problem.  For
scalar advection, the speed of each wave is simply the local advection speed at
the cell interface.  For non-linear problems and systems, an approximate
Riemann solver, such as a Roe solver \cite{roesolver}, is typically used.  Since
much of the physics of an application can be contained in the Riemann solver,
\forestclaw adopts \clawpack's interface to Riemann solvers effectively allowing
problems solved with \clawpack to be solvable in \forestclaw.  Wave limiters
are used to supress spurious oscillations resulting from truncation error in a typical
second order method.
% at the cost of a localized order reduction.
% \comment{DAC: Limiters formally reduce the order of accuracy.}
% wauld be an awkward senctence if order reduction is mentioned
\ignore{In order to
achieve second order accuracy wave limiters are used to suppress spurious
oscillations at sharp gradients in the flow.}

Data exchanges between neighboring patches are done via layers of
ghost cells extending the dimensions along the edges of each patch.
The interior edge values of a given patch overlap the ghost cell
region of a neighboring patch.  For the second order wave propagation
algorithm, two layers of ghost cells are sufficient.  This implies that
one layer of ghost patches is sufficient for $m \ge 4$.  Neighboring
patches at the same level of refinement simply copy their interior
edge values into a neighbors' ghost cells.  Neighboring fine grid
patches average their interior edge data to a coarser neighbor's ghost
cell values.  And neighboring coarse grid patches interpolate data
from their interior edge cells to their fine grid neighbor's ghost
cell values.  To avoid loss of conservation and the creation of
spurious extrema, we use a standard conservative, limited
interpolation scheme to interpolate values from the coarse grid to
fine grid coarse cells \cite{amrclaw, chombo}.  When sub-cycling, time accurate
data between coarse grids is used to fill in ghost cells for fine grids.
As we detail in the following section, this procedure can be extended
transparently to distributed parallelism by defining an abstract exchange
routine for ghost patch data.

@@

\begin{figure}
\begin{center}
\includegraphics[width=.4\columnwidth]{../figs_misc/forest_stacked}
\hspace{.05\columnwidth}
\begin{tikzpicture}[every node/.style={anchor=south west,inner sep=0pt},
                    x=1mm, y=1mm]
  \node at (0,0) {
    \includegraphics[width=.4\columnwidth]{../figs_misc/cubed3}
  };
  \node [fill=white, inner sep=2pt] at (28, 10) {1};
  \node [fill=white, inner sep=2pt] at (5, 27) {2};
  \node [fill=white, inner sep=2pt] at (41, 27) {2};
  \node [fill=white, inner sep=2pt] at (28, 36) {3};
\end{tikzpicture}
%\includegraphics[width=.4\columnwidth]{../figs_misc/cubed3}
\end{center}
\caption{Left: Forest of two quadtrees, partitioned among three MPI processes.
Each quadtree has its own coordinate orientation.  The situation in 3D
(octrees) is analogous.  Right: The leaves in a forest of six quadtrees that
serves as the computational domain for the cubed sphere.  An adhoc refinement
pattern has been 2:1 balanced in terms of neighbor sizes and partitioned
between five MPI processes (the three middle ones out of $0 \ldots 4$
are shown with a color scale from black to white).}
\label{fig:cubed3}
\end{figure}
